---
title: "GoFigr Clean Room Demo"
output: html_document
---

```{r setup, include=FALSE}
library(gofigR)
library(ggplot2)

knitr::opts_chunk$set(echo = TRUE)
gofigR::enable(url="http://localhost:9090",
               workspace_name="Scratchpad",
               analysis_name="Clean room in R",
               api_key="49cb3c49129a13d74908cbe631c47c673b3753191e71a3147ddbf48e611b3d51")
```

## Clean Room: Interactive Reproducibility

The `reproducible()` function captures everything needed to re-run a figure in
an isolated environment: source code, parameter values with widget metadata,
package versions, and input data. Call `publish()` inside the function body to
publish with full clean room metadata, enabling interactive parameter exploration
in the web viewer.

## Basic example: Iris scatter plot

Parameters are declared as function defaults using `slider()`, `dropdown()`,
`checkbox()`, and `text_input()`. Data frames use `static()`.

```{r iris_scatter, fig.width=8, fig.height=6}
reproducible(function(
    point_size = slider(2, min = 0.5, max = 5, step = 0.5),
    alpha = slider(0.7, min = 0.1, max = 1.0, step = 0.1),
    color_by = dropdown("Species", choices = c("Species", "Sepal.Width", "Petal.Width")),
    show_smooth = checkbox(FALSE),
    data = static(iris)
  ) {
    p <- ggplot(data, aes(x = Sepal.Length, y = Petal.Length, color = .data[[color_by]])) +
      geom_point(size = point_size, alpha = alpha) +
      labs(title = "Iris: Sepal Length vs Petal Length",
           x = "Sepal Length (cm)", y = "Petal Length (cm)") +
      theme_minimal()

    if (show_smooth) {
      p <- p + geom_smooth(method = "lm", se = FALSE)
    }

    publish(p, figure_name = "Iris Scatter")
  },
  packages = c("ggplot2")
)
```

## Histogram with configurable binning

```{r histogram, fig.width=8, fig.height=5}
reproducible(function(
    bins = slider(20L, min = 5L, max = 50L, step = 5L),
    fill_color = dropdown("steelblue", choices = c("steelblue", "coral", "forestgreen", "goldenrod")),
    show_density = checkbox(TRUE),
    title = text_input("Distribution of MPG"),
    data = static(mtcars)
  ) {
    p <- ggplot(data, aes(x = mpg)) +
      geom_histogram(bins = bins, fill = fill_color, color = "white", alpha = 0.8) +
      labs(title = title, x = "Miles per Gallon", y = "Count") +
      theme_minimal()

    if (show_density) {
      p <- p + geom_density(aes(y = after_stat(count)), linewidth = 0.8)
    }

    publish(p, figure_name = "MPG Distribution")
  },
  packages = c("ggplot2")
)
```

## Faceted analysis

```{r faceted, fig.width=10, fig.height=6}
reproducible(function(
    metric = dropdown("Sepal.Length", choices = c("Sepal.Length", "Sepal.Width",
                                                   "Petal.Length", "Petal.Width")),
    geom = dropdown("boxplot", choices = c("boxplot", "violin", "jitter")),
    data = static(iris)
  ) {
    p <- ggplot(data, aes(x = Species, y = .data[[metric]], fill = Species)) +
      theme_minimal() +
      labs(title = paste(metric, "by Species"), y = metric) +
      theme(legend.position = "none")

    if (geom == "boxplot") {
      p <- p + geom_boxplot(alpha = 0.7)
    } else if (geom == "violin") {
      p <- p + geom_violin(alpha = 0.7) + geom_jitter(width = 0.1, size = 0.8)
    } else {
      p <- p + geom_jitter(width = 0.2, size = 2, alpha = 0.6)
    }

    publish(p, figure_name = paste(metric, "by Species"))
  },
  packages = c("ggplot2")
)
```
