---
title: "TCGA-LUAD: Exploratory Analysis with gofigR"
output: html_document
---

```{r setup, include=FALSE}
library(gofigR)
library(SummarizedExperiment)
library(dplyr)
library(ggplot2)
library(tidyr)
library(stringr)
library(gridExtra)
library(ggplotify)

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

# Enable gofigR for this analysis. You can optionally pass an explicit analysis name.
gofigR::enable(TRUE, analysis_name = "TCGA-LUAD demo")
```

## Introduction

This document demonstrates how to use **gofigR** with a TCGA LUAD (lung adenocarcinoma) subset stored as a `SummarizedExperiment`.
We will:

- **Enable gofigR** and configure the session.
- **Load the LUAD expression dataset** from `TCGA-LUAD_TPM_SE_subset.rds`.
- **Tabulate patient demographics** and publish the table as a figure to GoFigr.
- **Run basic QC on genes** (mean and standard deviation of expression).
- **Explore gene expression by confounders**: gender, smoking status, and age.
- **Assess association of genes with sample type** (tumor vs normal) using boxplots and a table of effect sizes / p-values.

Throughout, all plots (and tabular displays rendered as plots) are published to GoFigr.

## Load TCGA-LUAD SummarizedExperiment

```{r load-data}
#se_path <- "TCGA-LUAD_TPM_SE_subset.rds"
se_path <- "../TCGA-LUAD_TPM_SE.rds"
stopifnot(file.exists(se_path))

luad_se <- readRDS(se_path)
luad_se
```

The object is a `SummarizedExperiment` with gene expression in the `assay` slot, gene annotations in `rowData`, and sample / patient metadata in `colData`.

For convenience we will extract:

```{r extract-components}
expr_mat <- as.matrix(assay(luad_se))
gene_info <- as.data.frame(rowData(luad_se))
sample_info <- as.data.frame(colData(luad_se))

head(sample_info)
```

## Transform expression to log10(1 + TPM)

We transform all expression values to log10(1 + TPM) to normalize the distribution and reduce the impact of highly expressed genes. This transformation is applied to the entire expression matrix before any downstream analysis.

```{r transform-expression}
# Transform expression to log10(1 + TPM)
expr_mat <- log10(1 + expr_mat)

# Also update the assay in the SummarizedExperiment object
assay(luad_se) <- expr_mat

message("Expression values transformed to log10(1 + TPM)")
message(paste("Expression range:", round(min(expr_mat, na.rm = TRUE), 3), 
              "to", round(max(expr_mat, na.rm = TRUE), 3)))
```

## Patient demographics

In this section we summarize patient-level demographics, using `sample_info`. The exact column names may vary depending on how TCGA data was prepared; here we use commonly present fields such as:

- **gender**
- **age_at_diagnosis** (which we convert to age in years, if present)
- **tobacco_smoking_history** or similar fields describing smoking status
- **sample_type** (e.g. Tumor vs Solid Tissue Normal)

```{r demographics-prep}
demo_df <- sample_info

# Try to standardize age in years if we have age_at_diagnosis (in days)
if ("age_at_diagnosis" %in% names(demo_df)) {
  demo_df$age_years <- demo_df$age_at_diagnosis / 365.25
}

# Try to identify a smoking variable (TCGA often encodes several variants)
smoking_cols <- grep("smok|tobacco", names(demo_df), ignore.case = TRUE, value = TRUE)
if (length(smoking_cols) > 0) {
  # Take the first smoking-related column for display
  demo_df$smoking_status <- as.character(demo_df[[smoking_cols[1]]])
}

# Keep a compact set of demographic variables (where present)
demo_vars <- c("patient", "bcr_patient_barcode", "gender", "age_years",
               "smoking_status", "sample_type")
demo_vars <- intersect(demo_vars, names(demo_df))

demographics <- demo_df %>%
  dplyr::select(dplyr::all_of(demo_vars))

head(demographics)
```

### Demographics summary table

We create a summary table of patient demographics and publish it to GoFigr. Since GoFigr doesn't support tables directly, we convert the table to a graphic using `gridExtra::tableGrob()` and `ggplotify::as.ggplot()`.

```{r demographics-table, fig.width=10, fig.height=6}
# Create a summary table showing first 20 rows of demographics
demo_table_display <- demographics %>%
  head(20) %>%
  dplyr::mutate_all(as.character) %>%
  dplyr::mutate_all(~ifelse(is.na(.), "", .))

# Convert table to graphic using gridExtra and ggplotify
demo_table_grob <- gridExtra::tableGrob(
  demo_table_display,
  theme = gridExtra::ttheme_minimal(
    base_size = 9,
    padding = unit(c(4, 4), "mm")
  )
)

demo_table_plot <- ggplotify::as.ggplot(demo_table_grob) +
  labs(title = "TCGA-LUAD Patient Demographics (first 20 samples)") +
  theme(plot.title = element_text(hjust = 0.5, size = 12, face = "bold"))

publish(demo_table_plot, data = demo_table_display,
        figure_name = "TCGA-LUAD demographics table")
```

We now compute simple counts / summaries by key variables and plot them as small bar charts to serve as graphical tables, which can be captured by gofigR.

### Gender distribution

```{r gender-table, fig.width=5, fig.height=4}
gender_counts <- demographics %>%
  dplyr::count(gender) %>%
  dplyr::mutate(pct = 100 * n / sum(n))

gender_plot <- ggplot(gender_counts, aes(x = gender, y = n, fill = gender)) +
  geom_col() +
  geom_text(aes(label = paste0(n, " (", round(pct, 1), "%)")),
            vjust = -0.2, size = 3) +
  labs(title = "Gender distribution",
       x = "Gender",
       y = "Number of patients") +
  theme_minimal() +
  theme(legend.position = "none")

publish(gender_plot, data = gender_counts, figure_name = "TCGA-LUAD gender distribution")
```

### Sample type distribution (tumor vs normal)

```{r sample-type-table, fig.width=5, fig.height=4}
if ("sample_type" %in% names(demographics)) {
  sample_type_counts <- demographics %>%
    dplyr::count(sample_type) %>%
    dplyr::mutate(pct = 100 * n / sum(n))

  sample_type_plot <- ggplot(sample_type_counts,
                             aes(x = sample_type, y = n, fill = sample_type)) +
    geom_col() +
    geom_text(aes(label = paste0(n, " (", round(pct, 1), "%)")),
              vjust = -0.2, size = 3) +
    labs(title = "Sample type distribution",
         x = "Sample type",
         y = "Number of samples") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 30, hjust = 1),
          legend.position = "none")

  publish(sample_type_plot, data = sample_type_counts,
          figure_name = "TCGA-LUAD sample type distribution")
} else {
  message("No sample_type column found in colData.")
}
```

### Smoking status (if available)

```{r smoking-table, fig.width=6, fig.height=4}
if ("smoking_status" %in% names(demographics)) {
  smoking_counts <- demographics %>%
    dplyr::count(smoking_status) %>%
    dplyr::mutate(pct = 100 * n / sum(n))

  smoking_plot <- ggplot(smoking_counts,
                         aes(x = smoking_status, y = n, fill = smoking_status)) +
    geom_col() +
    geom_text(aes(label = paste0(n, " (", round(pct, 1), "%)")),
              vjust = -0.2, size = 3) +
    labs(title = "Smoking status distribution",
         x = "Smoking status",
         y = "Number of samples / patients") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 30, hjust = 1),
          legend.position = "none")

  publish(smoking_plot, data = smoking_counts,
          figure_name = "TCGA-LUAD smoking status distribution")
} else {
  message("No smoking_status column inferred from colData.")
}
```

## Gene-level QC: mean and standard deviation

We now perform simple QC on gene expression across samples. For each gene we calculate:

- **Mean expression** across all samples.
- **Standard deviation (SD)** across all samples.

These summaries help identify genes with very low expression or little variability.

```{r gene-qc-summary}
gene_qc <- data.frame(
  gene_index = seq_len(nrow(expr_mat)),
  gene_id = if ("gene_id" %in% names(gene_info)) gene_info$gene_id else rownames(expr_mat),
  gene_name = if ("gene_name" %in% names(gene_info)) gene_info$gene_name else rownames(expr_mat),
  mean_expr = rowMeans(expr_mat, na.rm = TRUE),
  sd_expr = apply(expr_mat, 1, sd, na.rm = TRUE)
)

head(gene_qc)
```

### Gene QC summary table

We create a summary table showing the top and bottom genes by mean expression and standard deviation, and publish it to GoFigr.

```{r gene-qc-table, fig.width=12, fig.height=8}
# Create summary table with top and bottom genes
gene_qc_summary <- gene_qc %>%
  dplyr::arrange(desc(mean_expr)) %>%
  dplyr::slice_head(n = 5) %>%
  dplyr::mutate(rank = "Top 5 by mean") %>%
  dplyr::bind_rows(
    gene_qc %>%
      dplyr::arrange(mean_expr) %>%
      dplyr::slice_head(n = 5) %>%
      dplyr::mutate(rank = "Bottom 5 by mean")
  ) %>%
  dplyr::bind_rows(
    gene_qc %>%
      dplyr::arrange(desc(sd_expr)) %>%
      dplyr::slice_head(n = 5) %>%
      dplyr::mutate(rank = "Top 5 by SD")
  ) %>%
  dplyr::bind_rows(
    gene_qc %>%
      dplyr::arrange(sd_expr) %>%
      dplyr::slice_head(n = 5) %>%
      dplyr::mutate(rank = "Bottom 5 by SD")
  ) %>%
  dplyr::select(rank, gene_name, gene_id, 
                mean_expr_rounded = mean_expr, 
                sd_expr_rounded = sd_expr) %>%
  dplyr::mutate(
    mean_expr_rounded = round(mean_expr_rounded, 4),
    sd_expr_rounded = round(sd_expr_rounded, 4)
  ) %>%
  dplyr::mutate_all(as.character) %>%
  dplyr::rename(
    Rank = rank,
    `Gene Name` = gene_name,
    `Gene ID` = gene_id,
    `Mean Expression` = mean_expr_rounded,
    `SD Expression` = sd_expr_rounded
  )

# Convert table to graphic
gene_qc_table_grob <- gridExtra::tableGrob(
  gene_qc_summary,
  theme = gridExtra::ttheme_minimal(
    base_size = 9,
    padding = unit(c(4, 4), "mm")
  )
)

gene_qc_table_plot <- ggplotify::as.ggplot(gene_qc_table_grob) +
  labs(title = "Gene QC Summary: Top and Bottom Genes by Mean Expression and SD") +
  theme(plot.title = element_text(hjust = 0.5, size = 12, face = "bold"))

publish(gene_qc_table_plot, data = gene_qc_summary,
        figure_name = "TCGA-LUAD gene QC summary table")
```

### Distribution of mean expression

```{r mean-expression, fig.width=6, fig.height=4}
mean_expr_plot <- ggplot(gene_qc, aes(x = mean_expr)) +
  geom_histogram(bins = 30, color = "white", fill = "steelblue") +
  labs(title = "Distribution of mean expression across genes",
       x = "Mean expression",
       y = "Number of genes") +
  theme_minimal()

publish(mean_expr_plot, data = gene_qc,
        figure_name = "TCGA-LUAD gene mean expression distribution")
```

### Distribution of standard deviation

```{r sd-expression, fig.width=6, fig.height=4}
sd_expr_plot <- ggplot(gene_qc, aes(x = sd_expr)) +
  geom_histogram(bins = 30, color = "white", fill = "darkorange") +
  labs(title = "Distribution of expression standard deviation across genes",
       x = "Standard deviation of expression",
       y = "Number of genes") +
  theme_minimal()

publish(sd_expr_plot, data = gene_qc,
        figure_name = "TCGA-LUAD gene SD expression distribution")
```

### Mean vs standard deviation (gene-level scatter)

```{r mean-vs-sd, fig.width=6, fig.height=5}
mean_sd_scatter <- ggplot(gene_qc, aes(x = mean_expr, y = sd_expr)) +
  geom_point(alpha = 0.7, size = 1.5) +
  labs(title = "Mean vs SD of gene expression",
       x = "Mean expression",
       y = "Standard deviation") +
  theme_minimal()

publish(mean_sd_scatter, data = gene_qc,
        figure_name = "TCGA-LUAD mean vs SD of gene expression")
```

## Expression versus confounders

We next explore how gene expression varies with potential confounders:

- **Gender**
- **Smoking status** (if available)
- **Age**

We focus on genes of interest: **EGFR, TP53, ALK, ROS1, BRAF, MET**. These are key genes in lung adenocarcinoma and will be visualized in the plots below.

```{r long-gene-expression}
# Define genes of interest
genes_of_interest <- c("EGFR", "TP53", "ALK", "ROS1", "BRAF", "MET")

gene_long <- as.data.frame(t(expr_mat))
gene_long$sample_id <- rownames(gene_long)

gene_long <- gene_long %>%
  tidyr::pivot_longer(
    cols = -sample_id,
    names_to = "gene_row",
    values_to = "expression"
  )

# Attach gene names from rowData for nicer facet labels
if ("gene_name" %in% names(gene_info)) {
  gene_long$gene_name <- gene_info$gene_name[match(gene_long$gene_row, rownames(gene_info))]
} else {
  gene_long$gene_name <- gene_long$gene_row
}

# Filter to genes of interest only
gene_long <- gene_long %>%
  dplyr::filter(gene_name %in% genes_of_interest)

# Merge sample covariates
gene_long <- gene_long %>%
  dplyr::left_join(
    sample_info %>%
      dplyr::mutate(sample_id = rownames(sample_info)),
    by = "sample_id"
  )

# Verify we have the expected genes
message("Genes included in analysis: ", paste(unique(gene_long$gene_name), collapse = ", "))
head(gene_long)
```

### Expression by gender

```{r expr-by-gender, fig.width=8, fig.height=5}
if ("gender" %in% names(gene_long)) {
  expr_gender_plot <- ggplot(
    gene_long %>% dplyr::filter(!is.na(gender)),
    aes(x = gender, y = expression, fill = gender)
  ) +
    geom_boxplot(outlier.size = 0.7) +
    facet_wrap(~ gene_name, scales = "free_y") +
    labs(title = "Gene expression by gender",
         x = "Gender",
         y = "Expression") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 30, hjust = 1),
          legend.position = "none")

  publish(expr_gender_plot, data = gene_long,
          figure_name = "TCGA-LUAD gene expression by gender")
} else {
  message("No gender column available in colData.")
}
```

### Expression by smoking status (if available)

```{r expr-by-smoking, fig.width=10, fig.height=5}
if ("smoking_status" %in% names(gene_long)) {
  expr_smoking_plot <- ggplot(
    gene_long %>% dplyr::filter(!is.na(smoking_status)),
    aes(x = smoking_status, y = expression, fill = smoking_status)
  ) +
    geom_boxplot(outlier.size = 0.7) +
    facet_wrap(~ gene_name, scales = "free_y") +
    labs(title = "Gene expression by smoking status",
         x = "Smoking status",
         y = "Expression") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 30, hjust = 1),
          legend.position = "none")

  publish(expr_smoking_plot, data = gene_long,
          figure_name = "TCGA-LUAD gene expression by smoking status")
} else {
  message("No smoking_status variable available for expression plots.")
}
```

### Expression versus age

We next visualize the relationship between expression and age.

```{r expr-by-age, fig.width=10, fig.height=5}
age_col <- NULL
if ("age_years" %in% names(gene_long)) {
  age_col <- "age_years"
} else if ("age_at_diagnosis" %in% names(gene_long)) {
  age_col <- "age_at_diagnosis"
}

if (!is.null(age_col)) {
  expr_age_plot <- ggplot(
    gene_long %>% dplyr::filter(!is.na(.data[[age_col]])),
    aes(x = .data[[age_col]], y = expression)
  ) +
    geom_point(alpha = 0.6) +
    geom_smooth(method = "lm", se = TRUE, color = "steelblue") +
    facet_wrap(~ gene_name, scales = "free_y") +
    labs(
      title = paste0("Gene expression vs ", age_col),
      x = age_col,
      y = "Expression"
    ) +
    theme_minimal()

  publish(expr_age_plot, data = gene_long,
          figure_name = "TCGA-LUAD gene expression vs age")
} else {
  message("No age variable available for expression vs age plots.")
}
```

## Association of genes with sample type (tumor vs normal)

Finally, we analyze how gene expression differs by **sample type** (e.g. tumor vs normal). We:

1. **Fit linear models** of the form `expression ~ sample_type` for each gene.
2. Extract **effect sizes** and **p-values**.
3. Present the results as a table.
4. Create **boxplots** of expression by sample type for each gene.

```{r gene-sampletype-models}
if (!("sample_type" %in% names(gene_long))) {
  stop("sample_type column not found in colData; cannot compute tumor vs normal associations.")
}

gene_sampletype_results <- gene_long %>%
  dplyr::filter(!is.na(sample_type)) %>%
  dplyr::group_by(gene_name) %>%
  dplyr::do({
    df <- .
    # Simple linear model: expression ~ sample_type
    fit <- try(lm(expression ~ sample_type, data = df), silent = TRUE)
    if (inherits(fit, "try-error")) {
      return(data.frame(term = NA, estimate = NA, p.value = NA))
    }
    smry <- summary(fit)
    coefs <- as.data.frame(coef(smry))
    coefs$term <- rownames(coefs)
    coefs <- coefs %>%
      dplyr::select(term, estimate = Estimate, p.value = `Pr(>|t|)`)
    coefs
  }) %>%
  dplyr::ungroup()

gene_sampletype_results
```

The table below focuses on the `sample_type`-related terms in each model (i.e. the differences between non-baseline sample types and the reference).

```{r gene-sampletype-table, fig.width=10, fig.height=6}
sampletype_effects <- gene_sampletype_results %>%
  dplyr::filter(grepl("sample_type", term)) %>%
  dplyr::arrange(p.value) %>%
  dplyr::mutate(
    estimate_rounded = round(estimate, 4),
    p_value_formatted = formatC(p.value, format = "e", digits = 2),
    neg_log10_p = round(-log10(p.value), 2)
  )

# Create a formatted table for display
sampletype_table_display <- sampletype_effects %>%
  dplyr::select(gene_name, term, estimate_rounded, p_value_formatted, neg_log10_p) %>%
  dplyr::rename(
    Gene = gene_name,
    Term = term,
    Estimate = estimate_rounded,
    `P-value` = p_value_formatted,
    `-log10(P)` = neg_log10_p
  ) %>%
  dplyr::mutate_all(as.character)

# Convert table to graphic
sampletype_table_grob <- gridExtra::tableGrob(
  sampletype_table_display,
  theme = gridExtra::ttheme_minimal(
    base_size = 9,
    padding = unit(c(4, 4), "mm")
  )
)

sampletype_table_plot <- ggplotify::as.ggplot(sampletype_table_grob) +
  labs(title = "Gene Expression Association with Sample Type (Linear Model Results)") +
  theme(plot.title = element_text(hjust = 0.5, size = 12, face = "bold"))

publish(sampletype_table_plot, data = sampletype_table_display,
        figure_name = "TCGA-LUAD gene-sample type association table")

# Also create the bar plot visualization
effects_plot <- ggplot(sampletype_effects,
                       aes(x = reorder(paste(gene_name, term, sep = " : "), -log10(p.value)),
                           y = -log10(p.value))) +
  geom_col(fill = "darkred") +
  coord_flip() +
  labs(title = "Gene expression association with sample type",
       x = "Gene : sample_type term",
       y = "-log10(p-value)") +
  theme_minimal()

publish(effects_plot, data = sampletype_effects,
        figure_name = "TCGA-LUAD gene association with sample type")
```

### Boxplots of expression by sample type

```{r expr-boxplot-sampletype, fig.width=8, fig.height=5}
expr_sampletype_plot <- ggplot(
  gene_long %>% dplyr::filter(!is.na(sample_type)),
  aes(x = sample_type, y = expression, fill = sample_type)
) +
  geom_boxplot(outlier.size = 0.7) +
  facet_wrap(~ gene_name, scales = "free_y") +
  labs(title = "Gene expression by sample type (tumor vs normal)",
       x = "Sample type",
       y = "Expression") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1),
        legend.position = "none")

publish(expr_sampletype_plot, data = gene_long,
        figure_name = "TCGA-LUAD gene expression by sample type")
```

## Summary

In this short analysis we:

- **Enabled gofigR** for a TCGA LUAD analysis.
- **Loaded** a `SummarizedExperiment` from `TCGA-LUAD_TPM_SE_subset.rds`.
- **Summarized patient demographics** and published graphical tables to GoFigr.
- **Performed gene-level QC**, visualizing distributions of mean and SD and their joint relationship.
- **Explored expression versus confounders** (gender, smoking status, age).
- **Quantified the association between genes and sample type**, both via a model-based summary table and via boxplots.

Every figure, including the table-like graphics, has been **published to GoFigr**, enabling reproducible sharing and inspection of the underlying data and code.


