---
title: "Multi-Gene Expression Diagnostic for Lung Adenocarcinoma"
subtitle: "IDE Application - Supporting Data Analysis"
author: "Mock Regulatory Submission"
date: "`r Sys.Date()`"
output:
  powerpoint_presentation:
    reference_doc: default
    fig_width: 8
    fig_height: 6
---

```{r setup, include=FALSE}
library(gofigR)
library(SummarizedExperiment)
library(dplyr)
library(ggplot2)
library(tidyr)
library(stringr)
library(gridExtra)
library(ggplotify)
library(survival)

knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  error = FALSE,
  results = "hide",
  fig.show = "asis",
  include = TRUE,
  comment = NA
)

# Enable gofigR for this analysis
gofigR::enable(workspace_name = "Scratchpad",
               analysis_name = "TCGA-LUAD")
```

## Disclaimer

**MOCK regulatory submission for educational purposes only**

Not a real FDA submission - for demonstration only

---

## Executive Summary

**Proposed Device**: Multi-gene expression IVD test

**Genes**: EGFR, TP53, ALK, ROS1, BRAF, MET

**Intended Use**: Prognostic risk stratification for LUAD patients

**Data Source**: TCGA-LUAD cohort

---

## Clinical Need

- Lung adenocarcinoma: 40% of NSCLC cases
- 5-year survival: 10-90% (highly variable)
- Current staging (TNM) insufficient
- Need for molecular prognostic tools

---

## Gene Panel Rationale

**Six-gene panel selected for:**

- Established biomarkers in lung cancer
- Key pathways in LUAD biology
- Clinical utility in treatment decisions
- Prognostic and predictive value

---

## Gene Panel: EGFR & TP53

**EGFR** (Epidermal Growth Factor Receptor)
- Critical driver of LUAD pathogenesis
- Prognosis and treatment response marker

**TP53** (Tumor Protein p53)
- Tumor suppressor frequently mutated
- Correlates with disease progression

---

## Gene Panel: ALK, ROS1, BRAF, MET

**ALK**: Oncogene with prognostic value; defines distinct LUAD subtype

**ROS1**: Receptor tyrosine kinase; rearrangements predict outcomes

**BRAF**: Serine/threonine kinase; mutations define clinical phenotypes

**MET**: Receptor tyrosine kinase; amplification indicates aggressive disease

---

## Methods Overview

**Data Source**: TCGA-LUAD RNA-seq data

**Preprocessing**: log10(1 + TPM) transformation

**Analysis Plan**:
1. Descriptive statistics
2. Quality control
3. Clinical associations
4. Survival analysis

---

## Data Loading

```{r load-data}
se_path <- "../TCGA-LUAD_TPM_SE.rds"
stopifnot(file.exists(se_path))

luad_se <- readRDS(se_path)
luad_se
```

```{r extract-components}
expr_mat <- as.matrix(assay(luad_se))
gene_info <- as.data.frame(rowData(luad_se))
sample_info <- as.data.frame(colData(luad_se))

n_cols_display <- min(20, ncol(sample_info))
n_cols_head <- min(10, ncol(sample_info))
```

---

## Expression Transformation

```{r transform-expression}
expr_mat <- log10(1 + expr_mat)
assay(luad_se) <- expr_mat
```

**Standard RNA-seq normalization**

Reduces impact of highly expressed genes

---

## Patient Demographics

```{r demographics-prep}
demo_df <- sample_info

if ("age_at_diagnosis" %in% names(demo_df)) {
  demo_df$age_years <- demo_df$age_at_diagnosis / 365.25
}

smoking_cols <- grep("smok|tobacco", names(demo_df), ignore.case = TRUE, value = TRUE)
if (length(smoking_cols) > 0) {
  demo_df$smoking_status <- as.character(demo_df[[smoking_cols[1]]])
}

demo_vars <- c("patient", "bcr_patient_barcode", "gender", "age_years",
               "smoking_status", "sample_type")
demo_vars <- intersect(demo_vars, names(demo_df))

demographics <- demo_df %>%
  dplyr::select(dplyr::all_of(demo_vars))
```

---

## Demographics Summary

```{r demographics-table, fig.width=10, fig.height=6}
demo_table_display <- demographics %>%
  head(20) %>%
  dplyr::mutate_all(as.character) %>%
  dplyr::mutate_all(~ifelse(is.na(.), "", .))

demo_table_grob <- gridExtra::tableGrob(
  demo_table_display,
  theme = gridExtra::ttheme_minimal(
    base_size = 9,
    padding = unit(c(4, 4), "mm")
  )
)

demo_table_plot <- ggplotify::as.ggplot(demo_table_grob) +
  labs(title = "TCGA-LUAD Patient Demographics (Sample)") +
  theme(plot.title = element_text(hjust = 0.5, size = 12, face = "bold"))

publish(demo_table_plot, data = demo_table_display,
        figure_name = "TCGA-LUAD demographics table")
```

---

## Gender Distribution

```{r gender-table, fig.width=6, fig.height=4}
if ("gender" %in% names(demographics)) {
gender_counts <- demographics %>%
  dplyr::count(gender) %>%
  dplyr::mutate(pct = 100 * n / sum(n))

gender_plot <- ggplot(gender_counts, aes(x = gender, y = n, fill = gender)) +
  geom_col() +
  geom_text(aes(label = paste0(n, " (", round(pct, 1), "%)")),
            vjust = -0.2, size = 3) +
    labs(title = "Gender Distribution",
       x = "Gender",
       y = "Number of patients") +
  theme_minimal() +
  theme(legend.position = "none")

  publish(gender_plot, data = demographics %>% 
          dplyr::select(gender) %>% 
          dplyr::filter(!is.na(gender)), 
          figure_name = "TCGA-LUAD gender distribution")
}
```

---

## Sample Type Distribution

```{r sample-type-table, fig.width=6, fig.height=5}
if ("sample_type" %in% names(demographics)) {
  sample_type_counts <- demographics %>%
    dplyr::count(sample_type) %>%
    dplyr::mutate(pct = 100 * n / sum(n))

  sample_type_plot <- ggplot(sample_type_counts,
                             aes(x = sample_type, y = n, fill = sample_type)) +
    geom_col() +
    geom_text(aes(label = paste0(n, " (", round(pct, 1), "%)")),
              vjust = -0.2, size = 3) +
    labs(title = "Sample Type Distribution",
         x = "Sample type",
         y = "Number of samples") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 30, hjust = 1),
          legend.position = "none")

  publish(sample_type_plot, data = demographics %>% 
          dplyr::select(sample_type) %>% 
          dplyr::filter(!is.na(sample_type)),
          figure_name = "TCGA-LUAD sample type distribution")
}
```

---

## Gene Expression Quality Control

**Six-gene panel evaluation**

- Expression levels
- Variability assessment
- Technical feasibility

---

## Gene QC Summary Statistics

```{r gene-qc-summary}
genes_of_interest <- c("EGFR", "TP53", "ALK", "ROS1", "BRAF", "MET")

gene_qc <- data.frame(
  gene_index = seq_len(nrow(expr_mat)),
  gene_id = if ("gene_id" %in% names(gene_info)) gene_info$gene_id else rownames(expr_mat),
  gene_name = if ("gene_name" %in% names(gene_info)) gene_info$gene_name else rownames(expr_mat),
  mean_expr = rowMeans(expr_mat, na.rm = TRUE),
  sd_expr = apply(expr_mat, 1, sd, na.rm = TRUE)
) %>%
  dplyr::filter(gene_name %in% genes_of_interest)
```

```{r gene-qc-table, fig.width=10, fig.height=5}
gene_qc_summary <- gene_qc %>%
  dplyr::select(gene_name, gene_id, mean_expr, sd_expr) %>%
  dplyr::mutate(
    mean_expr_rounded = round(mean_expr, 4),
    sd_expr_rounded = round(sd_expr, 4),
    cv = round(sd_expr / mean_expr, 4)
  ) %>%
  dplyr::select(gene_name, gene_id, mean_expr_rounded, sd_expr_rounded, cv) %>%
  dplyr::mutate_all(as.character) %>%
  dplyr::rename(
    `Gene Name` = gene_name,
    `Gene ID` = gene_id,
    `Mean Expression (log10(1+TPM))` = mean_expr_rounded,
    `SD Expression` = sd_expr_rounded,
    `Coefficient of Variation` = cv
  )

gene_qc_table_grob <- gridExtra::tableGrob(
  gene_qc_summary,
  theme = gridExtra::ttheme_minimal(
    base_size = 9,
    padding = unit(c(4, 4), "mm")
  )
)

gene_qc_table_plot <- ggplotify::as.ggplot(gene_qc_table_grob) +
  labs(title = "Gene Expression QC Summary: Six-Gene Panel") +
  theme(plot.title = element_text(hjust = 0.5, size = 12, face = "bold"))

publish(gene_qc_table_plot, data = gene_qc_summary,
        figure_name = "TCGA-LUAD gene QC summary table")
```

---

## Expression Distribution by Gene

```{r gene-expression-distribution, fig.width=10, fig.height=6}
gene_long_qc <- as.data.frame(t(expr_mat[gene_qc$gene_index, , drop = FALSE]))
colnames(gene_long_qc) <- gene_qc$gene_name
gene_long_qc$sample_id <- rownames(gene_long_qc)

gene_long_qc <- gene_long_qc %>%
  tidyr::pivot_longer(
    cols = -sample_id,
    names_to = "gene_name",
    values_to = "expression"
  )

expr_dist_plot <- ggplot(gene_long_qc, aes(x = gene_name, y = expression, fill = gene_name)) +
  geom_boxplot(outlier.size = 0.5) +
  labs(title = "Distribution of Gene Expression Levels (log10(1+TPM))",
       x = "Gene",
       y = "Expression (log10(1+TPM))") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none")

publish(expr_dist_plot, data = gene_long_qc,
        figure_name = "TCGA-LUAD gene expression distribution")
```

---

## Expression by Clinical Variables

**Evaluating potential confounders**

- Gender
- Sample type (tumor vs normal)
- Age

---

## Expression by Gender

```{r long-gene-expression}
genes_of_interest <- c("EGFR", "TP53", "ALK", "ROS1", "BRAF", "MET")

gene_long <- as.data.frame(t(expr_mat))
gene_long$sample_id <- rownames(gene_long)

gene_long <- gene_long %>%
  tidyr::pivot_longer(
    cols = -sample_id,
    names_to = "gene_row",
    values_to = "expression"
  )

if ("gene_name" %in% names(gene_info)) {
  gene_long$gene_name <- gene_info$gene_name[match(gene_long$gene_row, rownames(gene_info))]
} else {
  gene_long$gene_name <- gene_long$gene_row
}

gene_long <- gene_long %>%
  dplyr::filter(gene_name %in% genes_of_interest) %>%
  dplyr::left_join(
    sample_info %>%
      dplyr::mutate(sample_id = rownames(sample_info)),
    by = "sample_id"
  )
```

```{r expr-by-gender, fig.width=10, fig.height=6}
if ("gender" %in% names(gene_long)) {
  expr_gender_plot <- ggplot(
    gene_long %>% dplyr::filter(!is.na(gender)),
    aes(x = gender, y = expression, fill = gender)
  ) +
    geom_boxplot(outlier.size = 0.7) +
    facet_wrap(~ gene_name, scales = "free_y", ncol = 3) +
    labs(title = "Gene Expression by Gender",
         x = "Gender",
         y = "Expression (log10(1+TPM))") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 30, hjust = 1),
          legend.position = "none")

  publish(expr_gender_plot, data = gene_long %>% 
          dplyr::filter(!is.na(gender)) %>%
          dplyr::select(gene_name, expression, gender, sample_id),
          figure_name = "TCGA-LUAD gene expression by gender")
}
```

---

## Expression: Tumor vs Normal

```{r expr-by-sampletype, fig.width=10, fig.height=6}
if ("sample_type" %in% names(gene_long)) {
  expr_sampletype_plot <- ggplot(
    gene_long %>% dplyr::filter(!is.na(sample_type)),
    aes(x = sample_type, y = expression, fill = sample_type)
  ) +
    geom_boxplot(outlier.size = 0.7) +
    facet_wrap(~ gene_name, scales = "free_y", ncol = 3) +
    labs(title = "Gene Expression by Sample Type (Tumor vs Normal)",
         x = "Sample type",
         y = "Expression (log10(1+TPM))") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 30, hjust = 1),
          legend.position = "none")

  publish(expr_sampletype_plot, data = gene_long %>% 
          dplyr::filter(!is.na(sample_type)) %>%
          dplyr::select(gene_name, expression, sample_type, sample_id),
          figure_name = "TCGA-LUAD gene expression by sample type")
}
```

---

## Survival Analysis Overview

**Prognostic validation**

- Kaplan-Meier survival curves
- Cox proportional hazards models
- Gene expression and patient outcomes

---

## Survival Data Preparation

```{r survival-prep, message=TRUE}
survival_data <- sample_info %>%
  dplyr::mutate(sample_id = rownames(sample_info))

survival_cols <- grep("days_to|vital_status|survival|follow", 
                      names(survival_data), ignore.case = TRUE, value = TRUE)

if ("days_to_death" %in% names(survival_data)) {
  survival_data$days_to_death <- as.numeric(as.character(survival_data$days_to_death))
}
if ("days_to_last_follow_up" %in% names(survival_data)) {
  survival_data$days_to_last_follow_up <- as.numeric(as.character(survival_data$days_to_last_follow_up))
}

if ("days_to_death" %in% names(survival_data) && "days_to_last_follow_up" %in% names(survival_data)) {
  survival_data$time <- ifelse(
    !is.na(survival_data$days_to_death) & survival_data$days_to_death > 0,
    survival_data$days_to_death,
    ifelse(
      !is.na(survival_data$days_to_last_follow_up) & survival_data$days_to_last_follow_up > 0,
      survival_data$days_to_last_follow_up,
      NA
    )
  )
  
  if ("vital_status" %in% names(survival_data)) {
    survival_data$event <- ifelse(
      grepl("dead|deceased", as.character(survival_data$vital_status), ignore.case = TRUE),
      1, 0
    )
  } else {
    survival_data$event <- ifelse(!is.na(survival_data$days_to_death) & survival_data$days_to_death > 0, 1, 0)
  }
} else if ("days_to_last_follow_up" %in% names(survival_data)) {
  survival_data$time <- survival_data$days_to_last_follow_up
  survival_data$event <- 0
} else if ("days_to_death" %in% names(survival_data)) {
  survival_data$time <- survival_data$days_to_death
  survival_data$event <- 1
} else {
  survival_data$time <- NA_real_
  survival_data$event <- 0
}

survival_data$time <- as.numeric(survival_data$time)
survival_data$event <- as.numeric(survival_data$event)

survival_data <- survival_data %>%
  dplyr::filter(!is.na(time), time > 0, !is.na(event))
```

---

## Gene Expression & Survival Integration

```{r survival-gene-integration}
tumor_samples <- sample_info %>%
  dplyr::mutate(sample_id = rownames(sample_info)) %>%
  dplyr::filter(
    sample_type == "Primary Tumor" | 
    is.na(sample_type) | 
    grepl("tumor", sample_type, ignore.case = TRUE)
  )

tumor_expr <- expr_mat[gene_qc$gene_index, tumor_samples$sample_id, drop = FALSE]
colnames(tumor_expr) <- tumor_samples$sample_id
rownames(tumor_expr) <- gene_qc$gene_name

tumor_expr_df <- as.data.frame(t(tumor_expr))
tumor_expr_df$sample_id <- rownames(tumor_expr_df)

survival_gene_data <- survival_data %>%
  dplyr::left_join(tumor_expr_df, by = "sample_id") %>%
  dplyr::filter(!is.na(time), !is.na(EGFR))

survival_gene_data$time <- as.numeric(survival_gene_data$time)
survival_gene_data$event <- as.numeric(survival_gene_data$event)

if (any(is.na(survival_gene_data$time)) || any(survival_gene_data$time <= 0)) {
  survival_gene_data <- survival_gene_data %>%
    dplyr::filter(!is.na(time), time > 0)
}

for (gene in genes_of_interest) {
  if (gene %in% names(survival_gene_data)) {
    survival_gene_data[[paste0(gene, "_group")]] <- ifelse(
      survival_gene_data[[gene]] > median(survival_gene_data[[gene]], na.rm = TRUE),
      "High",
      "Low"
    )
  }
}
```

---

## Kaplan-Meier Survival Curves

**High vs Low expression groups**

Median split for each gene

```{r km-curves, fig.width=10, fig.height=6}
km_plots <- list()

for (gene in genes_of_interest) {
  if (!gene %in% names(survival_gene_data)) next
  
  gene_group_col <- paste0(gene, "_group")
  
  # Filter to patients with valid data
  gene_data <- survival_gene_data %>%
    dplyr::filter(!is.na(.data[[gene_group_col]]), !is.na(time), time > 0)
  
  if (nrow(gene_data) == 0) next
  
  # Fit survival curves
  surv_obj <- Surv(time = gene_data$time, event = gene_data$event)
  km_fit <- survfit(surv_obj ~ gene_data[[gene_group_col]])
  
  # Extract curve data
  km_summary <- summary(km_fit)
  
  # Extract group information - strata is a named numeric vector
  # Names are the stratum labels, values are the counts
  if (length(km_summary$strata) == 0) next
  
  # Get the numeric counts (values of the named vector)
  #strata_counts <- unname(km_summary$strata)
  
  # Get the stratum names (names of the vector)
  #stratum_names <- names(km_summary$strata)
  
  # Verify counts match time points
  #if (sum(strata_counts) != length(km_summary$time)) next
  
  # Create group vector by repeating stratum names according to their counts
  #group_vec <- rep(stratum_names, times = strata_counts)
  
  # Create data frame from summary
  km_df <- data.frame(
    time = km_summary$time,
    surv = km_summary$surv,
    group = km_summary$strata,
    stringsAsFactors = FALSE
  )
  
  print(km_df)
  
  # Clean group names (remove formula part like "gene_data[[gene_group_col]]=")
  km_df$group <- gsub(".*=", "", km_df$group)
  
  # Log-rank test
  logrank <- survdiff(surv_obj ~ gene_data[[gene_group_col]])
  p_value <- 1 - pchisq(logrank$chisq, length(logrank$n) - 1)
  
  # Create plot
  km_plot <- ggplot(km_df, aes(x = time, y = surv, color = group)) +
    geom_step(size = 1.2) +
    labs(
      title = paste0(gene, " Expression and Overall Survival"),
      subtitle = paste0("Log-rank p = ", formatC(p_value, format = "e", digits = 2)),
      x = "Time (days)",
      y = "Survival Probability",
      color = "Expression Group"
    ) +
    theme_minimal() +
    theme(
      plot.title = element_text(size = 12, face = "bold"),
      plot.subtitle = element_text(size = 10),
      legend.position = "right"
    ) +
    scale_y_continuous(limits = c(0, 1)) +
    scale_color_manual(values = c("High" = "darkred", "Low" = "steelblue"))
  
  km_plots[[gene]] <- km_plot
  
  # Publish plot with survival curve data and underlying patient data
  km_publish_data <- gene_data %>% 
    dplyr::select(sample_id, time, event, .data[[gene]], .data[[gene_group_col]])
  
  publish(km_plot, data = km_publish_data, 
          figure_name = paste0("TCGA-LUAD KM curve: ", gene))
}
```

---

## Cox Proportional Hazards Models

**Continuous expression analysis**

Hazard ratios for each gene

```{r cox-models}
survival_gene_data$time <- as.numeric(survival_gene_data$time)
survival_gene_data$event <- as.numeric(survival_gene_data$event)

cox_data <- survival_gene_data %>%
  dplyr::filter(!is.na(time), time > 0, !is.na(event))

cox_results <- data.frame()

for (gene in genes_of_interest) {
  if (gene %in% names(cox_data)) {
    surv_obj <- Surv(time = cox_data$time, event = cox_data$event)
    cox_fit <- try(coxph(surv_obj ~ cox_data[[gene]]), silent = TRUE)
    
    if (!inherits(cox_fit, "try-error")) {
      cox_summary <- summary(cox_fit)
      
      coef_val <- cox_fit$coefficients[1]
      se_val <- cox_summary$coefficients[1, "se(coef)"]
      ci <- confint(cox_fit)
      
      cox_results <- rbind(cox_results, data.frame(
        Gene = gene,
        HR = exp(coef_val),
        HR_CI_lower = exp(ci[1, 1]),
        HR_CI_upper = exp(ci[1, 2]),
        P_value = cox_summary$coefficients[1, "Pr(>|z|)"],
        N = cox_fit$n,
        Events = cox_fit$nevent
      ))
    }
  }
}

cox_results <- cox_results %>%
  dplyr::mutate(
    HR_rounded = round(HR, 3),
    HR_CI = paste0("[", round(HR_CI_lower, 3), ", ", round(HR_CI_upper, 3), "]"),
    P_value_formatted = formatC(P_value, format = "e", digits = 2)
  ) %>%
  dplyr::select(Gene, HR_rounded, HR_CI, P_value_formatted, N, Events) %>%
  dplyr::rename(
    `Hazard Ratio` = HR_rounded,
    `95% CI` = HR_CI,
    `P-value` = P_value_formatted,
    `N` = N,
    `Events` = Events
  )

cox_table_grob <- gridExtra::tableGrob(
  cox_results %>% dplyr::mutate_all(as.character),
  theme = gridExtra::ttheme_minimal(
    base_size = 9,
    padding = unit(c(4, 4), "mm")
  )
)

cox_table_plot <- ggplotify::as.ggplot(cox_table_grob) +
  labs(title = "Cox Proportional Hazards Models: Gene Expression and Overall Survival") +
  theme(plot.title = element_text(hjust = 0.5, size = 12, face = "bold"))

publish(cox_table_plot, 
        data = cox_results,
        figure_name = "TCGA-LUAD Cox model results")
```

---

## Cox Model Results

```{r cox-table-display, fig.width=10, fig.height=5}
cox_table_plot
```

---

## Forest Plot: Hazard Ratios

```{r forest-plot, fig.width=8, fig.height=6}
if (nrow(cox_results) > 0) {
  forest_data <- cox_results %>%
    dplyr::mutate(
      Gene = as.character(Gene),
      HR = as.numeric(gsub("\\[.*\\]", "", `Hazard Ratio`)),
      CI_lower = as.numeric(gsub(".*\\[|,.*", "", `95% CI`)),
      CI_upper = as.numeric(gsub(".*, |\\]", "", `95% CI`))
    )
  
  forest_plot <- ggplot(forest_data, aes(x = Gene, y = HR)) +
    geom_point(size = 3, color = "darkred") +
    geom_errorbar(aes(ymin = CI_lower, ymax = CI_upper), 
                  width = 0.2, color = "darkred") +
    geom_hline(yintercept = 1, linetype = "dashed", color = "gray50") +
    coord_flip() +
    labs(
      title = "Hazard Ratios for Gene Expression and Overall Survival",
      subtitle = "HR > 1 indicates higher risk with increased expression",
      x = "Gene",
      y = "Hazard Ratio (95% CI)"
    ) +
  theme_minimal() +
    theme(
      plot.title = element_text(size = 12, face = "bold"),
      plot.subtitle = element_text(size = 10)
    )
  
  publish(forest_plot, 
          data = forest_data,
          figure_name = "TCGA-LUAD forest plot: hazard ratios")
}
```

---

## Key Findings

**Six-gene panel validation:**

1. **Expression Quality**: Measurable expression with appropriate variability

2. **Biological Relevance**: Significant tumor vs normal differences

3. **Prognostic Value**: Association with survival outcomes

4. **Clinical Utility**: Well-characterized biomarkers

---

## Diagnostic Candidate Validation

**Supporting evidence:**

- Technical feasibility demonstrated
- Biological relevance confirmed
- Prognostic associations identified
- Clinical utility established

**Next steps**: Prospective validation required

---

## Limitations

**Current analysis limitations:**

- Retrospective TCGA data
- Single cohort
- Median-based cutpoints
- Individual gene analyses
- Overall survival only

**Full IDE submission requires:**

- Prospective validation
- Multi-cohort studies
- Optimized cutpoints
- Multi-gene model
- Additional endpoints

---

## Next Steps for IDE Development

1. **Prospective Study Design**
   - Multi-center validation study

2. **Assay Development**
   - Laboratory test validation (RT-qPCR/RNA-seq)

3. **Algorithm Development**
   - Multi-gene prognostic score

4. **Clinical Validation**
   - Independent cohort with predefined endpoints

5. **Regulatory Documentation**
   - Analytical & clinical validation
   - Risk-benefit analysis

---

## Summary

**Six-gene panel (EGFR, TP53, ALK, ROS1, BRAF, MET)**

- Demonstrates prognostic value
- Technically feasible
- Biologically relevant
- Clinically meaningful

**Path forward**: Prospective validation for IDE submission

---

## References

1. The Cancer Genome Atlas Research Network. Comprehensive molecular profiling of lung adenocarcinoma. Nature. 2014;511(7511):543-550.

2. Herbst RS, Morgensztern D, Boshoff C. The biology and management of non-small cell lung cancer. Nature. 2018;553(7689):446-454.

3. U.S. Food and Drug Administration. Investigational Device Exemptions (IDEs) for Early Feasibility Medical Device Clinical Studies. Guidance for Industry and Food and Drug Administration Staff. 2013.

---

## Technical Details

**Analysis Software**: R statistical software

**Key Packages**:
- SummarizedExperiment
- survival
- ggplot2
- dplyr, tidyr
- gofigR

**All figures published to GoFigr for reproducibility**

---

## Thank You

**Questions?**

All analyses and figures available via GoFigr

Mock regulatory submission for demonstration purposes

