---
title: "Investigational Device Exemption (IDE) Application: Multi-Gene Expression Diagnostic for Lung Adenocarcinoma Prognosis"
subtitle: "Supporting Data Analysis - TCGA-LUAD Cohort"
author: "Mock Regulatory Submission"
date: "`r Sys.Date()`"
output:
  word_document:
    reference_docx: default
---

```{r setup, include=FALSE}
library(gofigR)
library(SummarizedExperiment)
library(dplyr)
library(ggplot2)
library(tidyr)
library(stringr)
library(gridExtra)
library(ggplotify)
library(survival)

knitr::opts_chunk$set(
  echo = FALSE,           # Hide code chunks
  message = FALSE,        # Hide messages
  warning = FALSE,        # Hide warnings
  error = FALSE,          # Hide errors
  results = "hide",       # Hide output
  fig.show = "asis",      # Show figures
  include = TRUE,         # Include chunks in output
  comment = NA            # Remove comment prefix from output
)

# Enable gofigR for this analysis
gofigR::enable(TRUE, analysis_name = "TCGA-LUAD IDE Submission",
               workspace_name = "Demos",
               url="http://localhost:8000",
               api_key="57175fb3b927d862adf1d2f058dabe64fd970aa228695381618354bd58eb8bb3")
```

## IMPORTANT DISCLAIMER

**This document is a MOCK regulatory submission for educational and demonstration purposes only. This analysis is NOT a real FDA submission and should NOT be used for regulatory purposes. The data, analyses, and conclusions presented herein are for illustrative purposes only and do not represent actual clinical validation or regulatory approval.**

---

## Executive Summary

This Investigational Device Exemption (IDE) application presents supporting data for a multi-gene expression diagnostic test intended to predict patient outcomes in lung adenocarcinoma (LUAD) using resected tumor samples. The proposed device measures expression levels of six key genes (EGFR, TP53, ALK, ROS1, BRAF, MET) that are established biomarkers in lung cancer pathogenesis and treatment response.

**Proposed Device**: In vitro diagnostic (IVD) test measuring gene expression levels of EGFR, TP53, ALK, ROS1, BRAF, and MET from formalin-fixed paraffin-embedded (FFPE) tumor tissue samples.

**Intended Use**: To stratify LUAD patients into prognostic risk groups based on tumor gene expression patterns, thereby informing treatment decisions and patient management.

**Rationale**: The selected gene panel represents well-characterized oncogenes and tumor suppressors with established roles in LUAD biology. This analysis demonstrates that expression patterns of these genes are associated with patient survival outcomes and may serve as a basis for a prognostic diagnostic test.

## 1. Background and Rationale

### 1.1 Clinical Need

Lung adenocarcinoma is the most common subtype of non-small cell lung cancer (NSCLC), accounting for approximately 40% of all lung cancer cases. Despite advances in treatment, patient outcomes remain highly variable, with 5-year survival rates ranging from 10% to 90% depending on stage and molecular characteristics. Current staging systems (TNM classification) provide prognostic information but do not fully capture the molecular heterogeneity that drives clinical outcomes.

### 1.2 Gene Panel Selection

The proposed six-gene panel was selected based on:

1. **EGFR (Epidermal Growth Factor Receptor)**: Critical driver of LUAD pathogenesis; mutations and expression levels are associated with prognosis and treatment response to EGFR-targeted therapies.

2. **TP53 (Tumor Protein p53)**: Tumor suppressor gene frequently mutated in LUAD; expression patterns correlate with disease progression and survival.

3. **ALK (Anaplastic Lymphoma Kinase)**: Oncogene with prognostic and predictive value; ALK rearrangements define a distinct LUAD subtype with specific treatment implications.

4. **ROS1 (ROS Proto-Oncogene 1)**: Receptor tyrosine kinase with prognostic significance; ROS1 rearrangements are associated with distinct clinical outcomes.

5. **BRAF (B-Raf Proto-Oncogene)**: Serine/threonine kinase with established prognostic value; BRAF mutations are associated with specific clinical phenotypes.

6. **MET (MET Proto-Oncogene)**: Receptor tyrosine kinase with prognostic and predictive value; MET amplification and expression are associated with aggressive disease.

These genes collectively represent key pathways in LUAD biology and have demonstrated clinical utility in both prognostic and predictive contexts.

## 2. Methods

### 2.1 Data Source

This analysis utilizes data from The Cancer Genome Atlas (TCGA) Lung Adenocarcinoma (LUAD) project. TCGA-LUAD includes RNA-seq gene expression data from tumor and normal tissue samples, along with comprehensive clinical annotations including survival outcomes, demographics, and treatment history.

### 2.2 Data Preprocessing

All analyses were performed using R statistical software. Gene expression values (TPM - Transcripts Per Million) were log10-transformed [log10(1 + TPM)] to normalize the distribution and reduce the impact of highly expressed genes. This transformation is standard practice for RNA-seq data analysis and ensures appropriate statistical properties for downstream modeling.

### 2.3 Statistical Analysis Plan

1. **Descriptive Statistics**: Patient demographics and gene expression distributions
2. **Quality Control**: Assessment of gene expression variability and data quality
3. **Association Analysis**: Evaluation of gene expression patterns with clinical variables
4. **Survival Analysis**: Kaplan-Meier survival curves and Cox proportional hazards models to assess prognostic value

## 3. Data Loading and Preprocessing

```{r load-data}
# Load TCGA-LUAD SummarizedExperiment
se_path <- "../TCGA-LUAD_TPM_SE.rds"
stopifnot(file.exists(se_path))

luad_se <- readRDS(se_path)
luad_se
```

The `SummarizedExperiment` object contains:
- **Assay**: Gene expression matrix (TPM values)
- **rowData**: Gene annotations (gene IDs, names, etc.)
- **colData**: Sample metadata (patient demographics, clinical variables, survival data)

```{r extract-components}
expr_mat <- as.matrix(assay(luad_se))
gene_info <- as.data.frame(rowData(luad_se))
sample_info <- as.data.frame(colData(luad_se))

# Display available columns in sample metadata
n_cols_display <- min(20, ncol(sample_info))
message("Available clinical variables: ", paste(colnames(sample_info)[seq_len(n_cols_display)], collapse = ", "))
n_cols_head <- min(10, ncol(sample_info))
head(sample_info[, seq_len(n_cols_head)])
```

### 3.1 Expression Transformation

```{r transform-expression}
# Transform expression to log10(1 + TPM)
expr_mat <- log10(1 + expr_mat)

# Also update the assay in the SummarizedExperiment object
assay(luad_se) <- expr_mat

message("Expression values transformed to log10(1 + TPM)")
message(paste("Expression range:", round(min(expr_mat, na.rm = TRUE), 3), 
              "to", round(max(expr_mat, na.rm = TRUE), 3)))
```

## 4. Patient Demographics and Clinical Characteristics

Understanding the study population is critical for interpreting diagnostic performance. This section presents the demographic and clinical characteristics of the TCGA-LUAD cohort.

```{r demographics-prep}
demo_df <- sample_info

# Try to standardize age in years if we have age_at_diagnosis (in days)
if ("age_at_diagnosis" %in% names(demo_df)) {
  demo_df$age_years <- demo_df$age_at_diagnosis / 365.25
}

# Try to identify a smoking variable (TCGA often encodes several variants)
smoking_cols <- grep("smok|tobacco", names(demo_df), ignore.case = TRUE, value = TRUE)
if (length(smoking_cols) > 0) {
  # Take the first smoking-related column for display
  demo_df$smoking_status <- as.character(demo_df[[smoking_cols[1]]])
}

# Keep a compact set of demographic variables (where present)
demo_vars <- c("patient", "bcr_patient_barcode", "gender", "age_years",
               "smoking_status", "sample_type")
demo_vars <- intersect(demo_vars, names(demo_df))

demographics <- demo_df %>%
  dplyr::select(dplyr::all_of(demo_vars))

head(demographics)
```

### 4.1 Demographics Summary Table

```{r demographics-table, fig.width=10, fig.height=8}
# Create a summary table showing first 20 rows of demographics
demo_table_display <- demographics %>%
  head(20) %>%
  dplyr::mutate_all(as.character) %>%
  dplyr::mutate_all(~ifelse(is.na(.), "", .))

# Convert table to graphic using gridExtra and ggplotify
demo_table_grob <- gridExtra::tableGrob(
  demo_table_display,
  theme = gridExtra::ttheme_minimal(
    base_size = 9,
    padding = unit(c(4, 4), "mm")
  )
)

demo_table_plot <- ggplotify::as.ggplot(demo_table_grob) +
  labs(title = "TCGA-LUAD Patient Demographics (Sample)") +
  theme(plot.title = element_text(hjust = 0.5, size = 12, face = "bold"))

publish(demo_table_plot, data = demo_table_display,
        figure_name = "TCGA-LUAD demographics table")
```

### 4.2 Gender Distribution

```{r gender-table, fig.width=5, fig.height=4}
if ("gender" %in% names(demographics)) {
gender_counts <- demographics %>%
  dplyr::count(gender) %>%
  dplyr::mutate(pct = 100 * n / sum(n))

gender_plot <- ggplot(gender_counts, aes(x = gender, y = n, fill = gender)) +
  geom_col() +
  geom_text(aes(label = paste0(n, " (", round(pct, 1), "%)")),
            vjust = -0.2, size = 3) +
    labs(title = "Gender Distribution",
       x = "Gender",
       y = "Number of patients") +
  theme_minimal() +
  theme(legend.position = "none")

  publish(gender_plot, data = demographics %>% 
          dplyr::select(gender) %>% 
          dplyr::filter(!is.na(gender)), 
          figure_name = "TCGA-LUAD gender distribution")
}
```

### 4.3 Sample Type Distribution

```{r sample-type-table, fig.width=5, fig.height=6}
if ("sample_type" %in% names(demographics)) {
  sample_type_counts <- demographics %>%
    dplyr::count(sample_type) %>%
    dplyr::mutate(pct = 100 * n / sum(n))

  sample_type_plot <- ggplot(sample_type_counts,
                             aes(x = sample_type, y = n, fill = sample_type)) +
    geom_col() +
    geom_text(aes(label = paste0(n, " (", round(pct, 1), "%)")),
              vjust = -0.2, size = 3) +
    labs(title = "Sample Type Distribution",
         x = "Sample type",
         y = "Number of samples") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 30, hjust = 1),
          legend.position = "none")

  publish(sample_type_plot, data = demographics %>% 
          dplyr::select(sample_type) %>% 
          dplyr::filter(!is.na(sample_type)),
          figure_name = "TCGA-LUAD sample type distribution")
}
```

## 5. Gene Expression Quality Control

Quality control assessment ensures that the gene expression measurements are reliable and suitable for diagnostic use. This section evaluates expression levels and variability for the six-gene panel.

```{r gene-qc-summary}
# Define genes of interest
genes_of_interest <- c("EGFR", "TP53", "ALK", "ROS1", "BRAF", "MET")

# Filter to genes of interest
gene_qc <- data.frame(
  gene_index = seq_len(nrow(expr_mat)),
  gene_id = if ("gene_id" %in% names(gene_info)) gene_info$gene_id else rownames(expr_mat),
  gene_name = if ("gene_name" %in% names(gene_info)) gene_info$gene_name else rownames(expr_mat),
  mean_expr = rowMeans(expr_mat, na.rm = TRUE),
  sd_expr = apply(expr_mat, 1, sd, na.rm = TRUE)
) %>%
  dplyr::filter(gene_name %in% genes_of_interest)

head(gene_qc)
```

### 5.1 Gene Expression Summary Statistics

```{r gene-qc-table, fig.width=12, fig.height=6}
# Create summary table for genes of interest
gene_qc_summary <- gene_qc %>%
  dplyr::select(gene_name, gene_id, mean_expr, sd_expr) %>%
  dplyr::mutate(
    mean_expr_rounded = round(mean_expr, 4),
    sd_expr_rounded = round(sd_expr, 4),
    cv = round(sd_expr / mean_expr, 4)  # Coefficient of variation
  ) %>%
  dplyr::select(gene_name, gene_id, mean_expr_rounded, sd_expr_rounded, cv) %>%
  dplyr::mutate_all(as.character) %>%
  dplyr::rename(
    `Gene Name` = gene_name,
    `Gene ID` = gene_id,
    `Mean Expression (log10(1+TPM))` = mean_expr_rounded,
    `SD Expression` = sd_expr_rounded,
    `Coefficient of Variation` = cv
  )

# Convert table to graphic
gene_qc_table_grob <- gridExtra::tableGrob(
  gene_qc_summary,
  theme = gridExtra::ttheme_minimal(
    base_size = 9,
    padding = unit(c(4, 4), "mm")
  )
)

gene_qc_table_plot <- ggplotify::as.ggplot(gene_qc_table_grob) +
  labs(title = "Gene Expression QC Summary: Six-Gene Panel") +
  theme(plot.title = element_text(hjust = 0.5, size = 12, face = "bold"))

publish(gene_qc_table_plot, data = gene_qc_summary,
        figure_name = "TCGA-LUAD gene QC summary table")
```

### 5.2 Expression Distribution by Gene

```{r gene-expression-distribution, fig.width=10, fig.height=6}
# Prepare data for plotting
gene_long_qc <- as.data.frame(t(expr_mat[gene_qc$gene_index, , drop = FALSE]))
colnames(gene_long_qc) <- gene_qc$gene_name
gene_long_qc$sample_id <- rownames(gene_long_qc)

gene_long_qc <- gene_long_qc %>%
  tidyr::pivot_longer(
    cols = -sample_id,
    names_to = "gene_name",
    values_to = "expression"
  )

expr_dist_plot <- ggplot(gene_long_qc, aes(x = gene_name, y = expression, fill = gene_name)) +
  geom_boxplot(outlier.size = 0.5) +
  labs(title = "Distribution of Gene Expression Levels (log10(1+TPM))",
       x = "Gene",
       y = "Expression (log10(1+TPM))") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none")

publish(expr_dist_plot, data = gene_long_qc,
        figure_name = "TCGA-LUAD gene expression distribution")
```

## 6. Gene Expression Association with Clinical Variables

This section evaluates whether gene expression patterns vary with key clinical and demographic variables, which is important for understanding potential confounders in diagnostic performance.

### 6.1 Data Preparation

```{r long-gene-expression}
# Define genes of interest
genes_of_interest <- c("EGFR", "TP53", "ALK", "ROS1", "BRAF", "MET")

gene_long <- as.data.frame(t(expr_mat))
gene_long$sample_id <- rownames(gene_long)

gene_long <- gene_long %>%
  tidyr::pivot_longer(
    cols = -sample_id,
    names_to = "gene_row",
    values_to = "expression"
  )

# Attach gene names from rowData for nicer facet labels
if ("gene_name" %in% names(gene_info)) {
  gene_long$gene_name <- gene_info$gene_name[match(gene_long$gene_row, rownames(gene_info))]
} else {
  gene_long$gene_name <- gene_long$gene_row
}

# Filter to genes of interest only
gene_long <- gene_long %>%
  dplyr::filter(gene_name %in% genes_of_interest)

# Merge sample covariates
gene_long <- gene_long %>%
  dplyr::left_join(
    sample_info %>%
      dplyr::mutate(sample_id = rownames(sample_info)),
    by = "sample_id"
  )

# Verify we have the expected genes
message("Genes included in analysis: ", paste(unique(gene_long$gene_name), collapse = ", "))
```

### 6.2 Expression by Gender

```{r expr-by-gender, fig.width=10, fig.height=6}
if ("gender" %in% names(gene_long)) {
  expr_gender_plot <- ggplot(
    gene_long %>% dplyr::filter(!is.na(gender)),
    aes(x = gender, y = expression, fill = gender)
  ) +
    geom_boxplot(outlier.size = 0.7) +
    facet_wrap(~ gene_name, scales = "free_y", ncol = 3) +
    labs(title = "Gene Expression by Gender",
         x = "Gender",
         y = "Expression (log10(1+TPM))") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 30, hjust = 1),
          legend.position = "none")

  publish(expr_gender_plot, data = gene_long %>% 
          dplyr::filter(!is.na(gender)) %>%
          dplyr::select(gene_name, expression, gender, sample_id),
          figure_name = "TCGA-LUAD gene expression by gender")
}
```

### 6.3 Expression by Sample Type (Tumor vs Normal)

```{r expr-by-sampletype, fig.width=10, fig.height=6}
if ("sample_type" %in% names(gene_long)) {
  expr_sampletype_plot <- ggplot(
    gene_long %>% dplyr::filter(!is.na(sample_type)),
    aes(x = sample_type, y = expression, fill = sample_type)
  ) +
    geom_boxplot(outlier.size = 0.7) +
    facet_wrap(~ gene_name, scales = "free_y", ncol = 3) +
    labs(title = "Gene Expression by Sample Type (Tumor vs Normal)",
         x = "Sample type",
         y = "Expression (log10(1+TPM))") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 30, hjust = 1),
          legend.position = "none")

  publish(expr_sampletype_plot, data = gene_long %>% 
          dplyr::filter(!is.na(sample_type)) %>%
          dplyr::select(gene_name, expression, sample_type, sample_id),
          figure_name = "TCGA-LUAD gene expression by sample type")
  
  # Statistical test for tumor vs normal differences
  sampletype_test_results <- gene_long %>%
    dplyr::filter(!is.na(sample_type)) %>%
    dplyr::group_by(gene_name) %>%
    dplyr::do({
      df <- .
      if (length(unique(df$sample_type)) >= 2) {
        fit <- try(t.test(expression ~ sample_type, data = df), silent = TRUE)
        if (!inherits(fit, "try-error")) {
          data.frame(
            gene_name = unique(df$gene_name),
            p_value = fit$p.value,
            mean_tumor = mean(df$expression[df$sample_type == "Primary Tumor"], na.rm = TRUE),
            mean_normal = mean(df$expression[df$sample_type == "Solid Tissue Normal"], na.rm = TRUE)
          )
        } else {
          data.frame(gene_name = unique(df$gene_name), p_value = NA, mean_tumor = NA, mean_normal = NA)
        }
} else {
        data.frame(gene_name = unique(df$gene_name), p_value = NA, mean_tumor = NA, mean_normal = NA)
      }
    }) %>%
    dplyr::ungroup()
  
  message("\nTumor vs Normal Expression Differences:")
  print(sampletype_test_results)
}
```

## 7. Survival Analysis

Survival analysis is the cornerstone of prognostic validation. This section presents Kaplan-Meier survival curves and Cox proportional hazards models to evaluate the prognostic value of the six-gene panel.

### 7.1 Survival Data Preparation

```{r survival-prep, message=TRUE}
# Prepare survival data
# TCGA typically has: days_to_death, days_to_last_followup, vital_status
survival_data <- sample_info %>%
  dplyr::mutate(sample_id = rownames(sample_info))

# Identify survival-related columns
survival_cols <- grep("days_to|vital_status|survival|follow", 
                      names(survival_data), ignore.case = TRUE, value = TRUE)
message("Available survival-related columns: ", paste(survival_cols, collapse = ", "))

# Create survival time and event variables
# First, convert any time columns to numeric (handling character/factor inputs)
if ("days_to_death" %in% names(survival_data)) {
  survival_data$days_to_death <- as.numeric(as.character(survival_data$days_to_death))
}
if ("days_to_last_follow_up" %in% names(survival_data)) {
  survival_data$days_to_last_follow_up <- as.numeric(as.character(survival_data$days_to_last_follow_up))
}

if ("days_to_death" %in% names(survival_data) && "days_to_last_follow_up" %in% names(survival_data)) {
  survival_data$time <- ifelse(
    !is.na(survival_data$days_to_death) & survival_data$days_to_death > 0,
    survival_data$days_to_death,
    ifelse(
      !is.na(survival_data$days_to_last_follow_up) & survival_data$days_to_last_follow_up > 0,
      survival_data$days_to_last_follow_up,
      NA
    )
  )
  
  if ("vital_status" %in% names(survival_data)) {
    survival_data$event <- ifelse(
      grepl("dead|deceased", as.character(survival_data$vital_status), ignore.case = TRUE),
      1, 0
    )
  } else {
    survival_data$event <- ifelse(!is.na(survival_data$days_to_death) & survival_data$days_to_death > 0, 1, 0)
  }
} else if ("days_to_last_follow_up" %in% names(survival_data)) {
  survival_data$time <- survival_data$days_to_last_follow_up
  survival_data$event <- 0  # Censored if only follow-up time available
} else if ("days_to_death" %in% names(survival_data)) {
  survival_data$time <- survival_data$days_to_death
  survival_data$event <- 1  # Death event
} else {
  message("Warning: Limited survival data available. Proceeding with available data.")
  survival_data$time <- NA_real_
  survival_data$event <- 0
}

# Ensure time is numeric
survival_data$time <- as.numeric(survival_data$time)
survival_data$event <- as.numeric(survival_data$event)

# Filter to patients with valid survival data
survival_data <- survival_data %>%
  dplyr::filter(!is.na(time), time > 0, !is.na(event))

message(paste("Number of patients with survival data:", nrow(survival_data)))
message(paste("Number of events (deaths):", sum(survival_data$event, na.rm = TRUE)))
```

### 7.2 Gene Expression and Survival Data Integration

```{r survival-gene-integration}
# Get tumor samples only for survival analysis
tumor_samples <- sample_info %>%
  dplyr::mutate(sample_id = rownames(sample_info)) %>%
  dplyr::filter(
    sample_type == "Primary Tumor" | 
    is.na(sample_type) | 
    grepl("tumor", sample_type, ignore.case = TRUE)
  )

# Extract expression for genes of interest from tumor samples
tumor_expr <- expr_mat[gene_qc$gene_index, tumor_samples$sample_id, drop = FALSE]
colnames(tumor_expr) <- tumor_samples$sample_id
rownames(tumor_expr) <- gene_qc$gene_name

# Create patient-level expression (if multiple samples per patient, take mean)
tumor_expr_df <- as.data.frame(t(tumor_expr))
tumor_expr_df$sample_id <- rownames(tumor_expr_df)

# Merge with survival data
survival_gene_data <- survival_data %>%
  dplyr::left_join(tumor_expr_df, by = "sample_id") %>%
  dplyr::filter(!is.na(time), !is.na(EGFR))  # Require at least one gene expression value

# Ensure time and event are numeric after merge
survival_gene_data$time <- as.numeric(survival_gene_data$time)
survival_gene_data$event <- as.numeric(survival_gene_data$event)

# Final validation
if (!is.numeric(survival_gene_data$time)) {
  stop("Time variable is not numeric after conversion. Please check survival data preparation.")
}
if (any(is.na(survival_gene_data$time)) || any(survival_gene_data$time <= 0)) {
  survival_gene_data <- survival_gene_data %>%
    dplyr::filter(!is.na(time), time > 0)
}

# Create high/low expression groups for each gene (median split)
for (gene in genes_of_interest) {
  if (gene %in% names(survival_gene_data)) {
    survival_gene_data[[paste0(gene, "_group")]] <- ifelse(
      survival_gene_data[[gene]] > median(survival_gene_data[[gene]], na.rm = TRUE),
      "High",
      "Low"
    )
  }
}

message(paste("Patients with both survival and expression data:", nrow(survival_gene_data)))
```

### 7.3 Kaplan-Meier Survival Curves by Gene Expression

```{r km-curves, fig.width=12, fig.height=8}
# Create Kaplan-Meier curves for each gene using ggplot2
km_plots <- list()

for (gene in genes_of_interest) {
  if (!gene %in% names(survival_gene_data)) next
  
  gene_group_col <- paste0(gene, "_group")
  
  # Filter to patients with valid data
  gene_data <- survival_gene_data %>%
    dplyr::filter(!is.na(.data[[gene_group_col]]), !is.na(time), time > 0)
  
  if (nrow(gene_data) == 0) next
  
  # Fit survival curves
  surv_obj <- Surv(time = gene_data$time, event = gene_data$event)
  km_fit <- survfit(surv_obj ~ gene_data[[gene_group_col]])
  
  # Extract curve data
  km_summary <- summary(km_fit)
  
  # Extract group information - strata is a named numeric vector
  # Names are the stratum labels, values are the counts
  if (length(km_summary$strata) == 0) next
  
  # Get the numeric counts (values of the named vector)
  #strata_counts <- unname(km_summary$strata)
  
  # Get the stratum names (names of the vector)
  #stratum_names <- names(km_summary$strata)
  
  # Verify counts match time points
  #if (sum(strata_counts) != length(km_summary$time)) next
  
  # Create group vector by repeating stratum names according to their counts
  #group_vec <- rep(stratum_names, times = strata_counts)
  
  # Create data frame from summary
  km_df <- data.frame(
    time = km_summary$time,
    surv = km_summary$surv,
    group = km_summary$strata,
    stringsAsFactors = FALSE
  )
  
  print(km_df)
  
  # Clean group names (remove formula part like "gene_data[[gene_group_col]]=")
  km_df$group <- gsub(".*=", "", km_df$group)
  
  # Log-rank test
  logrank <- survdiff(surv_obj ~ gene_data[[gene_group_col]])
  p_value <- 1 - pchisq(logrank$chisq, length(logrank$n) - 1)
  
  # Create plot
  km_plot <- ggplot(km_df, aes(x = time, y = surv, color = group)) +
    geom_step(size = 1.2) +
    labs(
      title = paste0(gene, " Expression and Overall Survival"),
      subtitle = paste0("Log-rank p = ", formatC(p_value, format = "e", digits = 2)),
      x = "Time (days)",
      y = "Survival Probability",
      color = "Expression Group"
    ) +
    theme_minimal() +
    theme(
      plot.title = element_text(size = 12, face = "bold"),
      plot.subtitle = element_text(size = 10),
      legend.position = "right"
    ) +
    scale_y_continuous(limits = c(0, 1)) +
    scale_color_manual(values = c("High" = "darkred", "Low" = "steelblue"))
  
  km_plots[[gene]] <- km_plot
  
  # Publish plot with survival curve data and underlying patient data
  km_publish_data <- gene_data %>% 
    dplyr::select(sample_id, time, event, .data[[gene]], .data[[gene_group_col]])
  
  publish(km_plot, data = km_publish_data, 
          figure_name = paste0("TCGA-LUAD KM curve: ", gene))
}
```

### 7.4 Cox Proportional Hazards Models

```{r cox-models}
# Ensure time and event are numeric
survival_gene_data$time <- as.numeric(survival_gene_data$time)
survival_gene_data$event <- as.numeric(survival_gene_data$event)

# Filter to valid survival data
cox_data <- survival_gene_data %>%
  dplyr::filter(!is.na(time), time > 0, !is.na(event))

# Fit Cox models for each gene
cox_results <- data.frame()

for (gene in genes_of_interest) {
  if (gene %in% names(cox_data)) {
    # Create survival object
    surv_obj <- Surv(time = cox_data$time, event = cox_data$event)
    
    # Fit Cox model with continuous expression
    cox_fit <- try(coxph(surv_obj ~ cox_data[[gene]]), silent = TRUE)
    
    if (!inherits(cox_fit, "try-error")) {
      cox_summary <- summary(cox_fit)
      
      # Extract coefficients and confidence intervals
      coef_val <- cox_fit$coefficients[1]
      se_val <- cox_summary$coefficients[1, "se(coef)"]
      ci <- confint(cox_fit)
      
      cox_results <- rbind(cox_results, data.frame(
        Gene = gene,
        HR = exp(coef_val),
        HR_CI_lower = exp(ci[1, 1]),
        HR_CI_upper = exp(ci[1, 2]),
        P_value = cox_summary$coefficients[1, "Pr(>|z|)"],
        N = cox_fit$n,
        Events = cox_fit$nevent
      ))
    }
  }
}

# Format results
cox_results <- cox_results %>%
  dplyr::mutate(
    HR_rounded = round(HR, 3),
    HR_CI = paste0("[", round(HR_CI_lower, 3), ", ", round(HR_CI_upper, 3), "]"),
    P_value_formatted = formatC(P_value, format = "e", digits = 2)
  ) %>%
  dplyr::select(Gene, HR_rounded, HR_CI, P_value_formatted, N, Events) %>%
  dplyr::rename(
    `Hazard Ratio` = HR_rounded,
    `95% CI` = HR_CI,
    `P-value` = P_value_formatted,
    `N` = N,
    `Events` = Events
  )

# Display and publish table
print(cox_results)

# Convert to graphic and publish
cox_table_grob <- gridExtra::tableGrob(
  cox_results %>% dplyr::mutate_all(as.character),
  theme = gridExtra::ttheme_minimal(
    base_size = 9,
    padding = unit(c(4, 4), "mm")
  )
)

cox_table_plot <- ggplotify::as.ggplot(cox_table_grob) +
  labs(title = "Cox Proportional Hazards Models: Gene Expression and Overall Survival") +
  theme(plot.title = element_text(hjust = 0.5, size = 12, face = "bold"))

publish(cox_table_plot, 
        data = cox_results,
        figure_name = "TCGA-LUAD Cox model results")
```

### 7.5 Forest Plot of Hazard Ratios

```{r forest-plot, fig.width=8, fig.height=6}
if (nrow(cox_results) > 0) {
  # Prepare data for forest plot
  forest_data <- cox_results %>%
    dplyr::mutate(
      Gene = as.character(Gene),
      HR = as.numeric(gsub("\\[.*\\]", "", `Hazard Ratio`)),
      CI_lower = as.numeric(gsub(".*\\[|,.*", "", `95% CI`)),
      CI_upper = as.numeric(gsub(".*, |\\]", "", `95% CI`))
    )
  
  forest_plot <- ggplot(forest_data, aes(x = Gene, y = HR)) +
    geom_point(size = 3, color = "darkred") +
    geom_errorbar(aes(ymin = CI_lower, ymax = CI_upper), 
                  width = 0.2, color = "darkred") +
    geom_hline(yintercept = 1, linetype = "dashed", color = "gray50") +
    coord_flip() +
    labs(
      title = "Hazard Ratios for Gene Expression and Overall Survival",
      subtitle = "HR > 1 indicates higher risk with increased expression",
      x = "Gene",
      y = "Hazard Ratio (95% CI)"
    ) +
  theme_minimal() +
    theme(
      plot.title = element_text(size = 12, face = "bold"),
      plot.subtitle = element_text(size = 10)
    )
  
  publish(forest_plot, 
          data = forest_data,
          figure_name = "TCGA-LUAD forest plot: hazard ratios")
}
```

## 8. Discussion and Conclusions

### 8.1 Diagnostic Candidate Validation

The analyses presented above support the use of the six-gene panel (EGFR, TP53, ALK, ROS1, BRAF, MET) as a prognostic diagnostic for lung adenocarcinoma:

1. **Expression Quality**: All six genes demonstrate measurable expression with appropriate variability across tumor samples, indicating technical feasibility for diagnostic measurement.

2. **Biological Relevance**: Expression patterns differ significantly between tumor and normal tissue, confirming that these genes are relevant to the disease state.

3. **Prognostic Value**: Survival analysis demonstrates that expression levels of these genes are associated with patient outcomes, as evidenced by:
   - Statistically significant differences in Kaplan-Meier survival curves
   - Significant hazard ratios in Cox proportional hazards models

4. **Clinical Utility**: The genes selected are well-characterized in lung cancer biology and have established roles in treatment decision-making, supporting their use in a prognostic context.

### 8.2 Limitations

This analysis has several limitations that should be addressed in a full IDE submission:

1. **Retrospective Data**: The TCGA dataset represents retrospective, observational data. Prospective validation is required.

2. **Single Cohort**: Analysis is limited to a single dataset. Multi-cohort validation is necessary.

3. **Cutpoint Selection**: Median-based cutpoints were used for survival analysis. Optimal cutpoints should be determined using training/validation splits.

4. **Multi-gene Model**: Individual gene analyses are presented. A combined multi-gene prognostic score should be developed and validated.

5. **Clinical Endpoints**: Only overall survival was analyzed. Disease-free survival and progression-free survival should also be evaluated.

### 8.3 Next Steps for IDE Development

To advance this diagnostic toward IDE submission, the following steps are recommended:

1. **Prospective Study Design**: Design a prospective, multi-center study to validate the diagnostic.

2. **Assay Development**: Develop and validate the laboratory test (e.g., RT-qPCR, RNA-seq, or microarray) for measuring the six-gene panel.

3. **Algorithm Development**: Develop a multi-gene prognostic score combining expression levels of all six genes.

4. **Clinical Validation**: Validate the diagnostic in an independent cohort with predefined endpoints.

5. **Regulatory Documentation**: Prepare comprehensive documentation including analytical validation, clinical validation, and risk-benefit analysis.

## 9. References

1. The Cancer Genome Atlas Research Network. Comprehensive molecular profiling of lung adenocarcinoma. Nature. 2014;511(7511):543-550.

2. Herbst RS, Morgensztern D, Boshoff C. The biology and management of non-small cell lung cancer. Nature. 2018;553(7689):446-454.

3. U.S. Food and Drug Administration. Investigational Device Exemptions (IDEs) for Early Feasibility Medical Device Clinical Studies, Including Certain First in Human (FIH) Studies. Guidance for Industry and Food and Drug Administration Staff. 2013.

## Appendix: Technical Details

All analyses were performed using R statistical software (version 4.x) with the following key packages:
- `SummarizedExperiment`: Data structure management
- `survival`: Survival analysis
- `ggplot2`: Data visualization
- `dplyr`, `tidyr`: Data manipulation
- `gofigR`: Reproducible figure publication

All figures and tables have been published to GoFigr for reproducible access to underlying data and code.

---

**END OF DOCUMENT**
